---
title: "mplyr"
author: "Giuseppe Paleologo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

mplyr helps users manipulate `array` objects, by offering alignment, split-apply-reduce, join and accumulation operations. The syntax is similar to `dplyr`, hence the name.

## Converting from data frame to array
There is a method `as.array.data.frame` to perform a pivot on a data.frame. The second argument is the aggregating function, the third one is the variable to be aggregating variable, the ones afterward are the marginal variables.  
```{r}
library(mplyr)
A <- as.array(diamonds, mean, x, clarity, cut)
B  <- as.array(diamonds, mean, x, clarity, cut, color)
A
```
and there is a summary function for arrays.
...{r}
summary(A)
summary(B)
```
The array has *axes*. These are name assigned to each dimension. `table` objects in R already use them. They can be retrieved and set:
```{r}
library(mplyr)
A <- as.array(diamonds, sum, x, clarity, cut)
axes(A)
axes(A) <- toupper(axes(A))
A %<>% set_axes(tolower(axes(A)))
```
## Subsetting
Subsetting by a condition on labels works with the same syntax as dplyr:
```{r}
A %>% filter(clarity %in% c('I1', 'SI2'), cut %in% c('Good','Very Good'))
```
There is also SE version, which admits formulas and quoted arguments; but it should take an argument at a time. It is still undergoing some testing, so please test it yourself before using and report issues:
```{r}
A %>% 
  filter_(~clarity %in% c('I1', 'SI2')) %>% 
  filter_("cut %in% c('Good','Very Good')")
```
## split-apply-reduce
If we want to aggregate the array across one or more dimension, the `group_by` will first partition the data and the `aggregate` will aggregate them. `group_by` takes as arguments the original array and one or more functions of the argument that map to a set of values, typically of lower cardinality than the domain so that one or more values will be aggregated; and `group_by` takes an array (that has been grouped_by) and an aggregating function.
```{r}
A <- array(1:7300, dim=c(365,10,2), dimnames=list(date=format(as.Date('1970-01-01')+1:365), 
                                                  stock=toupper(letters[1:10]),
                                                  feature=paste0('Feat',1:2)))  
summary(A)
A %<>% 
  group_by(substr(date,1,7)) %>%
  aggregate(FUN=mean, na.rm=TRUE)
summary(A)
```
## Aligning
Oftentime, two or more arrays are not aligned in that, while they may have the same number of dimensions (say 2-D matrices, 3-D arrays, up to dimension 5), they may have different sizes and/or misaligned labels. The function `align_array(list(X1, X2), all.dim=c(T,F), na.value=NA)` performs an alignment on a list of two or more arrays/matrices/tables. The alignment can be outer (on the union of each axis' labels) or inner (on their intersection). The axis names for the input arrays must be identical, and the dimension order must be the same. If not, you can change the order with `aperm` and change the axis names with `axis<-` or `set_axis`.
```{r}
  # we use vadr for unpacking
  library(vadr)
  X0 <- structure(Titanic, class='array')
  X1 <- X0[1:2,,,]
  X2 <- X0[-(1:2),,,]
  X3 <- accumulate(list(X1,X2), all.dim = c(TRUE, TRUE, TRUE, TRUE), FUN =`+`, na.value = 0)
  # let's align axes first
  bind[X3, X0] <- align_array(list(X3, X0), all.dim=c(T,T,T,T))
  table(X0 == X3)
```
## Joining
There is a also a joining function. In the case of arrays, an inner join would not be well defined unless an operation to operate on elements with overlapping indices. This is equivalent to the composition of an alignment and an element-wise function, and we prefer allowing the user compose these two primitives. For arrays with non-overlapping items, we have a join_all
```{r}
  library(vadr)
  X0 <- structure(Titanic, class='array')
  X1 <- X0[1:2,,1,1]
  X2 <- X0[-(1:2),,1,1]

  accumulate(list(X1,X2), all.dim = c(TRUE, TRUE), FUN =`+`, na.value = 0)
  join_all(list(X1, X2))
```
